---
title: 'Class 19: Genome Informatics'
author: 'Jasmine Lee (PID: A15583527)'
date: "11/30/2021"
output: github_document
---

# Section 1. Proportion of G/G in a Population

Downloaded a CSV file from Ensemble <https://uswest.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel>.

Here we read this CSV file.

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```


Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G.

```{r}
table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

Let's now dig into this further.


# Section 4. Population Scale Analysis

Downloaded a TXT file with the normalization of 230 samples on a genome level <https://bioboot.github.io/bggn213_W19/classmaterial/
rs8067378_ENSG00000172057.6.txt>. We want to determine whether there is any association of the 4 asthma-associated SNPs on ORMDL3 expression. 


> **Q13**. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.
**Hint**: The **read.table()**, **summary()** and **boxplot()** functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the median value from saving the output of the **boxplot()** function to an R object and examining this object. There is also the **median()** and **summary()** function that you can use to check your understanding.
 
```{r}
# Read TXT file in to R and store as expr
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
 
```{r}
# Determine total number of samples
nrow(expr)
```

```{r}
# Determine the sample size for each genotype
table(expr$geno)
```

```{r}
# Create a boxplot of expression vs. genotype and store as pop
pop <- boxplot(exp~geno, data=expr)
```

```{r}
# Determine the median expression levels for each genotype
# The third row displays the median values for each genotype
round(pop$stats, 2)
```

*Answer*: The sample size is 108 for genotype A|A, 233 for genotype A|G, and 121 for genotype G|G. The median expression level is 31.25 for genotype A|A, 25.06 for genotype A|G, and 20.07 for genotype G|G.


> **Q14**. Generate a boxplot with a box per genotype. What could you infer from the relative expression value between A|A and G|G displayed in this plot? Does the SNP affect the expression of ORMDL3? 
**Hint**: An example boxplot is provided overleaf - yours does not need to be as polished as this one.

```{r}
library(ggplot2)
```

Let's make a boxplot.

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE) +
  xlab("Genotype") +
  ylab("Expression")
```

*Answer*: There appears to be a statistically significant difference in expression between the genotypes A|A and G|G. Thus, I infer that there is an association between the genotype of G|G at this location and reduced expression of ORMDL3. The SNP affects the expression of ORMDL3.