---
title: 'Class 10: Halloween Candy'
author: "Jasmine Lee (PID: A15583527)"
date: "10/28/2021"
output: pdf_document
---

# 1. Importing the data

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy <- read.csv(url, row.names=1)
head(candy, n=3)
```

> **Q1**. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

*Answer*: There are 85 different candy types.


> **Q2**. How many fruity candy types are in this dataset?

```{r}
sum(candy$fruity)
```

*Answer*: There are 38 fruity candy types.


# 2. What is your favorite candy?

> **Q3**. What is your favorite candy in the dataset, and what is its winpercent value?

```{r}
candy["Sour Patch Kids",]$winpercent
```
 *Answer*: My favorite candy is Sour Patch Kids, and its winpercent value is 59.864. 
 

> **Q4**. What is the winpercent value for "Kit Kat"?

```{r}
candy["Kit Kat",]$winpercent
```

*Answer*: The winpercent value for "Kit Kat" is 76.7686.


> **Q5**. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

*Answer*: The winpercent value for "Tootsie Roll Snack Bars" is 49.6535.


```{r}
# install.packages("skimr")
library("skimr")
skim(candy)
```

> **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

*Answer*: The 'winpercent' column appears to be on a different scale.


> **Q7**. What do you think a zero and one represent for the candy$chocolate column?

*Answer*: I think a zero represents that the candy is not chocolate whereas a one represents that the candy is made of chocolate.


> **Q8**. Plot a histogram of winpercent values.

```{r}
winpercent <- candy$winpercent
hist(winpercent)
```

 > **Q9**. Is the distribution of winpercent values symmetrical?

*Answer*: No, the distribution of winpercent values is not symmetrical. The data is skewed to the right.


> **Q10**. Is the center of the distribution above or below 50%?

*Answer*: The center of the distribution is below 50%.


> **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

First need to find all the chocolate candy rows in the 'candy' data.

```{r}
inds1 <- as.logical(candy$chocolate)
chocolate <- candy[inds1,]$winpercent
```

Do the same for fruity candy.

```{r}
inds2 <- as.logical(candy$fruity)
fruity <- candy[inds2,]$winpercent
```

Compare by finding the means.

```{r}
mean(chocolate)
mean(fruity)
```

*Answer*: On average, chocolate candy is higher ranked than fruit candy.


> **Q12**. Is this difference statistically significant?

```{r}
t.test(chocolate, fruity)
```

*Answer*: Yes, this difference is statistically significant.


# 3. Overall candy rankings

```{r}
# Sort the candy dataset by winpercent
head(candy[order(candy$winpercent),], n=5)
```

> **Q13**. What are the five least liked candy types in this set?

*Answer*: The five least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.


> **Q14**. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

*Answer*: The top 5 all time favorite candy types are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter cup.


> **Q15**. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col()
```


> *Q16**. This is quite ugly; use the reorder() function to get the bars sorted by winpercent.

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy), winpercent)) + geom_col()
```


```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "blue"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy), winpercent)) + geom_col(fill=my_cols)
```


> **Q17**. What is the worst ranked chocolate candy?

*Answer*: The worst ranked chocolate candy is Sixlets.


> **Q18**. What is the best ranked fruity candy?

*Answer*: The best ranked fruity candy is Starburst.


# 4. Taking a look at pricepercent

```{r}
# install.packages("ggrepel")
library(ggrepel)

# Plot of price vs. win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=2, max.overlaps = 35)
```

> **Q19**. Which candy type is the highest ranked in terms of winpercent for the least money- i.e. offers the most bang for your buck?

*Answer*: Reese's Miniatures is the highest ranked in terms of winpercent for the least money.


> **Q20**. What are the top 5 most expensive candy types in the dataset, and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing=TRUE)
head(candy[ord,c(11,12)], n=5)
```

*Answer*: The top five most expensive candy types are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. Of these, Nik L Nip is the least popular. 


> **Q21**. Make a barplot again with geom_col(), this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so-called "dot chart" or "lollipop" chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


# 5. Exploring the correlation structure

```{r}
# install.packages("corrplot")
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> **Q22**. Examining this plot, what two variables are anti-correlated (i.e. have minus values)?

*Answer*: Chocolate and fruity are anti-correlated.


> **Q23**. Similarly, what two variables are most positively correlated?

*Answer*: Each variable is most positively correlated to itself. For example, chocolate is most positively correlated to chocolate.


# 6. Principal component analysis

```{r}
pca <- prcomp(candy,scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
# Take a look at PCA loadings
par(mar=c(8, 4, 2, 2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> **Q24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

*Answer*: Fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. These make sense to me. 