---
title: 'Class 17: Vaccination Rate Mini-Project'
author: 'Jasmine Lee (PID: A15583527)'
date: "11/23/2021"
output: 
  pdf_document:
    latex_engine: xelatex
---

# Background

As we approach a period of travel and larger gatherings, let's have a look at vaccination rates across the State.

We will take data from the CA.gov site here: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code. 

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> **Q1**. What column details the total number of people fully vaccinated?

*Answer*: The column "persons fully vaccinated" details the total number of people fully vaccinated.


> **Q2**. What column details the ZIP code tabulation area?

*Answer*: The column "zip code tabulation area" details the ZIP code tabulation area.


> **Q3**. What is the earliest date in this dataset?

```{r}
head(vax$as_of_date)
```

*Answer*: The earliest date in this dataset is 2021-01-05. 


> **Q4**. What is the latest date in this dataset?

```{r}
tail(vax$as_of_date)
```

*Answer*: The latest date in this dataset is 2021-11-16.


Let's call the 'skim()' function from the **skimr** package to get a quick overview of this dataset.

```{r}
library(skimr)
skimr::skim(vax)
```

> **Q5**. How many numeric columns are in this dataset?

*Answer*: There are 9 numeric columns in this dataset.


> **Q6**. Note that there are "missing values" in the dataset. How many NA values are there in the persons_fully_vaccinated column?

*Answer*: There are 8,256 NA values in the person_fully_vaccinated column.


> **Q7**. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
sum(is.na(vax$persons_fully_vaccinated)) / nrow(vax)
```

*Answer*: 10.17% of persons_fully_vaccinated values are missing.


> **Q8**. [Optional] Why might this data be missing?

*Answer*: Optional.


# Working with Dates

We will use the **lubridate** package to make life a lot easier when dealing with dates and times.

```{r}
library(lubridate)
```

```{r}
today()
```

Here we make our 'as_of_date' column lubridate format... 

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now I can do useful math with dates more easily.

```{r}
today()-vax$as_of_date[1]
```

```{r}
vax$as_of_date[nrow(vax)]-vax$as_of_date[1]
```

> **Q9**. How many days have passed since the last update of the dataset?

```{r}
today()-vax$as_of_date[nrow(vax)]
```

*Answer*: It has been 7 days since the last entry.


> **Q10**. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

*Answer*: There are 46 unique dates in the dataset.


# Working with ZIP Codes

```{r}
library(zipcodeR)
```

```{r}
# Find centroid of La Jolla 92037 ZIP code area
geocode_zip('92037')
```

```{r}
# Calculate distance between centroids of any two ZIP codes
zip_distance('92037', '92109')
```

More usefully, we can pull census data about ZIP code areas (including median household income, etc.).

```{r}
reverse_zipcode(c('92037', '92109'))
```

We can use this reverse_zipcode() to pull census data later on for any or all ZIP code areas we might be interested in.


# Focus on San Diego Area

```{r}
table(vax$county)
```
We will subset with base R.

```{r}
inds <- vax$county=="San Diego"
head(vax[inds,])
```

But let's use the **dplyr** package and its **filter()** function.

```{r}
library(dplyr)

sd <- filter(vax, county=="San Diego")

nrow(sd)
```

Using **dplyr** is more convenient when we are subsetting across multiple criteria. For example:

```{r}
sd.10 <- filter(vax, county=="San Diego" & age5_plus_population>10000)
```

> **Q11**. How many distinct ZIP codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

*Answer*: There are 107 distinct ZIP codes listed for San Diego County.


> **Q12**. What San Diego County ZIP code area has the largest 12+ population in this dataset?

```{r}
which.max(sd$age12_plus_population)
sd$zip_code_tabulation_area[23]
```

*Answer*: The San Diego County ZIP code area of 92154 has the largest 12+ population in this dataset.


Using **dplyr**, select all San Diego "county" entries on "as_of_date" "2021-11-09".

```{r}
sd.11.09 <- filter(vax, county=="San Diego" & as_of_date=="2021-11-09")
```

> **Q13**. What is the overall average "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2021-11-09"?

```{r}
mean(sd.11.09$percent_of_population_fully_vaccinated, na.rm=TRUE)
```

*Answer*: The overall average "Percent of Population Fully Vaccinated" value is 67.27567%.


We can look at the 6-number summary.

```{r}
summary(sd.11.09$percent_of_population_fully_vaccinated)
```


> **Q14**. Using either ggplot or base R graphics, make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of "2021-11-09".

*Answer*:

Using base R plots

```{r}
hist(sd.11.09$percent_of_population_fully_vaccinated,
     main="Histogram of Vaccination Rates Across San Diego County",
     xlab="Percent Fully Vaccinated on 2021-11-09",
     ylab="Frequency")
```

Using ggplot

```{r}
library(ggplot2)

ggplot(sd.11.09) + 
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins=10) +
  labs(x="Percent Fully Vaccinated on 2021-11-09", y="Count (ZIP Code Areas",
       title="Histogram of Vaccination Rates Across San Diego County")
```


# Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> **Q15**. Using **ggplot**, make a graph of the vaccination rate time course for the 92037 ZIP code area.

*Answer*:

```{r}
ggplot(ucsd) +
  aes(as_of_date,percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", 
       title="Vaccination Rate for La Jolla CA 92037")
```


# Comparing 92037 to Other Similar Sized Areas?

Let's return to the full dataset and look across every ZIP code area with a population at least as large as that of 92037 on "2021-11-16".

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population>36144 & as_of_date=="2021-11-16")
head(vax.36)
```

> **Q16**. Calculate the mean "Percent of Population Fully Vaccinated" for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date "2021-11-16". Add this as a straight horizontal line to your plot from above with the geom_hline() function.

```{r}
vaccination.36 <- mean(vax.36$percent_of_population_fully_vaccinated)
```

*Answer*:

```{r}
ggplot(ucsd) +
  aes(as_of_date,percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", 
       title="Vaccination Rate for La Jolla CA 92037") +
  geom_hline(yintercept=vaccination.36, color="red", linetype="dashed")
  
```


> **Q17**. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

*Answer*:

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```


> **Q18**. Using ggplot, generate a histogram of this data.

*Answer*:

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins=40) +
  labs(x="Percent Vaccinated", y="Count")
```


> **Q19**. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date=="2021-11-16") %>%
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)

vax %>% filter(as_of_date=="2021-11-16") %>%
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
 
*Answer*: The 92109 ZIP code area is above the average value of 0.6630 I calculated for all these above. However, the 92040 ZIP code area is below the average value.


> **Q20**. Finally make a time course of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population>36144)

ggplot(vax.36.all) +
  aes(as_of_date, percent_of_population_fully_vaccinated, group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", 
       title="Vaccination Rate Across California",
       subtitle="Only areas with a population above 36k are shown.") +
  geom_hline(yintercept=0.66, linetype="dashed")
  
```


> **Q21*. How do you feel about traveling for Thanksgiving and meeting for in-person class next week?

*Answer*: Considering the trend of rising cases in the U.S. as well as other countries and the lower-than-expected vaccination rates analyzed through this activity, I feel apprehensive about traveling for Thanksgiving and meeting for in-person class next week.
